A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-05 latex/main.tex
Pass through disco, including limitations of study
Closes #917
Closes #725

\section*{DISCUSSION}
There are two major contributions of this work. The first is an
Ancestral Recombination Graph (ARG) for 2.48 million SARS-CoV-2 genomes.
Through multiple layers of validation against results from the 
literature, we have demonstrated that this ARG successfully synthesises
the evolutionary features of genetic inheritance, point mutations,
deletions and recombination in a single cohesive structure.
The ARG is freely available and easily accessible using the 
tskit software ecosystem, a rich and efficient platform for 
recombination-aware analyses widely used in population genomics.
This combination of a deeply validated ARG with a pre-existing
and highly capable software platform forms a unique resource.
Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
have helped spur the development of numerous ARG-aware 
methods in human 
genomics~\cite{Tagami2024tstrait,Mejia2025using} [NOTE add more refs 
if we like this].
We hope that the ARG provided here will similarly
enable a range of recombination-aware analyses of SARS-CoV-2, and 
catalyse a wave of methodological improvements which fully integrate
recombination into pathogen genomics.

@@

% SHZ: WIP
We close by emphasizing several current limitations of sc2ts
and suggest potential directions for improvement
which we hope will inspire future development of ARG inference
methods for pathogens beyond SARS-CoV2.

There are several ways to extend the HMM.
First, we currently treat multi-base deletions as missing data,
adding indels to the ARG \textit{post hoc}.
However, by explicitly modeling deletions as single HMM events, and
accounting for fact that reversion of deletions is more likely to
occur through recombination than back-mutation,
we should be able to recover additional signals of recombination
and reduce spurious recombinants.
Second, the current HMM assumes a simplistic model of equiprobable
mutation between bases with a uniform mutation rate along the genome.
However, genomic analyses have revealed
transition-to-transversion rate bias~\cite{Bloom2023,Simmonds2024} and
sequence context-dependent mutation rate variation
across the SARS-CoV-2 genome~\cite{Haddox2025}.
Using a more realistic mutation model and
accounting for mutation rate heterogeneity
could improve inference of local genetic relationships in the ARG.
Third, geography can be accounted for in the HMM,
favoring attachment of samples to geographically close lineages.
This should also improve precise identification of recombinant parents,
because parents sampled in the same region or geographically close regions
are expected to be more plausible
than parents sampled in geographically distant regions.

% TODO: Add a short discussion of the node dates and
% some references to the literature, closing issue 725.

We have developed and used filters
to detect time-traveling samples in the Viridian dataset and
to exclude them from the ARG,
yielding a set of samples with reliable collection dates.
Future studies could leverage the ARG
to flag wrongly dated samples or
to infer the dates of samples
with missing or incomplete collection dates in the metadata.

It is challenging to accurately reconstruct global events
when there are regions with poor sampling and
jurisdictions with long delays
before sequences become publicly
available~\cite{Lancet2021,Kalia2021,Chen2022}.
For example, revisiting the ARG topology periodically
(e.g., rewinding the ARG a month and repeating
with samples that became available later)
might improve the accuracy of the genealogy.

We have developed and used several methods
to exclude spurious recombination events \textit{post hoc},
including filtering by supporting loci per parent and
splitting of suspiciously long branches (e.g., Delta and Omicron BA.2).
These methods could instead be run during ARG inference,
immediately after sample matching,
reducing the number of ARG post-processing steps.

Finally, an inherent limitation of detecting recombination from sequence data,
which applies to all methods,
is that it is impossible to detect events with few (or no) supporting bases,
as these cannot be distinguished from mutation events.
This limitation makes it particularly challenging
to detect events near the termini of linear viral genomes.

>>>>

\section*{DISCUSSION}
There are two major contributions of this work. The first is an
Ancestral Recombination Graph (ARG) for 2.48 million SARS-CoV-2 genomes.
Through multiple layers of validation against results from the 
literature, we have demonstrated that this ARG successfully synthesises
the evolutionary features of genetic inheritance, point mutations,
deletions and recombination in a single cohesive structure.
The ARG is freely available and easily accessible using the 
tskit software ecosystem, a rich and efficient platform for 
recombination-aware analyses widely used in population genomics.
This combination of a deeply validated ARG with a pre-existing
and highly capable software platform forms a unique resource.
Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
have helped spur the development of numerous ARG-aware 
methods in human 
genomics~\cite{Mejia2025using,Grundler2025geographic,
Lehmann2025on,Terhorst2025accelerated}.
We hope that the ARG provided here will similarly
enable a range of recombination-aware analyses of SARS-CoV-2, and 
catalyse a wave of methodological improvements which fully integrate
recombination into pathogen genomics.
We envisage the ARG as a long-term stable resource---building on the 
high-quality, freely available reassemblies of
the Viridian dataset~\cite{Hunt2024}---to facilitate the study 
of the pandemic phase of SARS-CoV-2 in an open and reproducible manner.

@@

We have focused on applying sc2ts to the Viridian dataset rather 
than the larger GISAID resource~\cite{Shu2017} for two reasons.
Firstly, the Viridian dataset is of higher quality and major efforts
have been made to remove sources of error. Applying sc2ts to 
GISAID data is straightforward, but significant additional efforts 
would be required to deal with the various data quality issues.
Secondly, the Viridian data is free to download without restrictions,
enabling us to share the resulting ARG in the same manner.

% Need to be careful here to list stuff that sounds interesting and 
% hard enough to not get asked to do it.
The sc2ts method may be extended and improved in several directions.
Firstly, the simplistic parameterisation of the LS HMM used 
here could be extended to include more realistic
models with mutation rate variation along the genome~\cite{Haddox2025}
and transition-to-transversion rate bias~\cite{Bloom2023,Simmonds2024}.
Secondly, the current handling of deletions by post-hoc parsimony
could be improved either by a more nuanced data-masking strategy
(in which, e.g., all but the right-most base of known deletions is 
masked as missing data during HMM matching), or by the more 
challenging approach of incorporating multi-base events into the HMM directly.
Sc2ts is currently specialised for SARS-CoV-2 data, but the methods
may be readily adapted to other pathogens.

We have used the tsdate method~\cite{Wohns2022,Pope2025} to date the 
sc2ts ARG, and initial results are promising. However, dating 
of phylogenies in itself is a complex task~\cite{Kumar2016advances}
and methods for incorporating information from metadata of uncertain 
quality into large trees are in their 
early stages~\cite{sanderson2021chronumental} and there is certainly
substantial room for improvement.
Much of the complexity of sc2ts lies in 
dealing with ``time traveller'' samples, where sampling dates 
provided in metadata are substantially incorrect. 
The present ARG may provide 
a way of improving this, by iteratively matching samples against
it to estimate dates for all sequences, and adjusting the dates
for time-travellers detected in this way. Sc2ts could then
be run without the complex filters required to remove 
time-travellers and unconditionally include all but the most divergent
sequences. Other studies may also benefit from the 
improved sample dates.

[Final para is prelim. Acknowledge limitations, forward looking 
for how we might deal with them]
Finally, it is important to acknowledge that sc2ts performs best
with dense sampling, and the dataset is heavily geographically biased.
[cite ~\cite{Lancet2021,Kalia2021,Chen2022}?]
Long branches, cause particular problems. Future improvements would 
ideally detect when data is sparse and switch to more sophisticated
models, as discussed in~\cite{Bisschop2025likelihoods}.

% It is challenging to accurately reconstruct global events
% when there are regions with poor sampling and
% jurisdictions with long delays
% before sequences become publicly
% available~\cite{Lancet2021,Kalia2021,Chen2022}.
% For example, revisiting the ARG topology periodically
% (e.g., rewinding the ARG a month and repeating
% with samples that became available later)
% might improve the accuracy of the genealogy.

% We have developed and used several methods
% to exclude spurious recombination events \textit{post hoc},
% including filtering by supporting loci per parent and
% splitting of suspiciously long branches (e.g., Delta and Omicron BA.2).
% These methods could instead be run during ARG inference,
% immediately after sample matching,
% reducing the number of ARG post-processing steps.

% Finally, an inherent limitation of detecting recombination from sequence data,
% which applies to all methods,
% is that it is impossible to detect events with few (or no) supporting bases,
% as these cannot be distinguished from mutation events.
% This limitation makes it particularly challenging
% to detect events near the termini of linear viral genomes.
